---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
### Ian Arbuckle

## Loading and preprocessing the data

The data are in a tidy format. The only change for ease of processing will be to interpret the **date** variable as a POSIXct data type.

```{r}
a <- read.csv("activity.csv")
a$date <- as.POSIXct(a$date)
```

## What is mean total number of steps taken per day?

The **dplyr** functions for grouping and summarizing will be useful here. First, set up a new tibble with the sum of steps per day, excluding **NAs** per the assignment instructions:

```{r message=FALSE}
library(dplyr)
totala <- a %>%
  group_by(date) %>%
  summarize(total = sum(steps, na.rm = TRUE))
```

This dataset can be displayed now as a histogram.

```{r message=FALSE}
library(ggplot2)
p <- ggplot(totala, aes(total))
p + geom_histogram() + labs(x = "Number of Steps", y = "Days")
```

And a mean and median value can be determined:

```{r}
totala %>%
  summarize(Mean = mean(total), Median = median(total))
```

## What is the average daily activity pattern?

First, we need to group and average the number of steps across the 5-minute intervals.

```{r}
inta <- a %>%
  group_by(interval) %>%
  summarize(avg = mean(steps, na.rm = TRUE))
```

Then we can plot these data with the interval minutes on the x-axis. To represent the time series, first the **interval** variable should be interpreted as a 24 hr time of hours and minutes. 

```{r message=FALSE}
library(lubridate)
tstr <- mapply(function(x, y) paste0(rep(x, y), collapse = "")
               , 0
               , 4 - nchar(inta$interval))
inta$dttm <- paste0(tstr, inta$interval)
inta$dttm <- strptime(inta$dttm, format = "%H%M")
```

Then the average steps can be plotted. 

```{r}
p <- ggplot(inta, aes(dttm, avg))
p + geom_line() + labs(x = "Time", y = "Avg Steps") + scale_x_datetime(date_labels = "%H:%M")
```

With the averaged steps in the **inta** dataset, we can check which 5-minute interval has the largest count: 

```{r}
inta[which(inta$avg == max(inta$avg)), 1]
```

This corresponds to the 5-minute interval beginning at 8:35 which appears supported by the time-series graph above. 

## Imputing missing values

There are a couple of quick ways to obtain the number of **NA** values in the dataset. My favorite is just using the **summary** function:

```{r}
summary(a)
```

My strategy for imputing the values for these **NAs** will be to use the mean of the particular 5 minute interval, as stored in the *inta* frame.

```{r}
b <- merge(a, inta, by.x = 3, by.y = 1)
b$steps <- with(b, ifelse(is.na(b$steps), b$avg, b$steps))
b$dttm <- as.POSIXct(b$dttm)
```

With imputed values now present in the **b** dataset, the histogram of steps by day may show differences from that containing **NAs**. 

```{r}
totalb <- b %>%
  group_by(date) %>%
  summarize(total = sum(steps))
p <- ggplot(totalb, aes(total))
p + geom_histogram(binwidth = 600) + labs(x = "Number of Steps", y = "Days")
```

And the by-day mean and median values can be recomputed to show the impact of imputed values:

```{r}
totalb %>%
  summarize(Mean = mean(total), Median = median(total))
```

Imputing values for **NAs** in this case creates a distribution that has a more normal shape, and causes the mean and median to converge.

## Are there differences in activity patterns between weekdays and weekends?

Adding in a factor variable to the dataset with imputed values to classify days of activity as either a weekday or a weekend.

```{r}
b$day_type <- as.factor(with(b, ifelse(weekdays(b$date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")))
```

With a factor available to distinguish, we can now plot a comparison between weekday and weekend activity.

```{r}
intb <- b %>%
  group_by(day_type, interval) %>%
  summarize(avg = mean(steps))

tstr <- mapply(function(x, y) paste0(rep(x, y), collapse = "")
               , 0
               , 4 - nchar(intb$interval))

intb$dttm <- paste0(tstr, intb$interval)
intb$dttm <- strptime(intb$dttm, format = "%H%M")
intb$dttm <- as.POSIXct(intb$dttm)
p <- ggplot(intb, aes(dttm, avg))
p <- p + geom_line() + labs(x = "Time", y = "Avg Steps") + scale_x_datetime(date_labels = "%H:%M")
p + facet_wrap(~day_type, nrow = 2)
```

We can see that weekdays tend to have a higher early mode, but subsequent modes tend to fall below the corresponding intervals of the weekends.